Refactoring `nls()`: a constructive critique

### 2023-02-03 RefactoringNLS/JNAB230203.Rmd
add rest

>>>>

## Code rationalization and documentation

When we started *Improvements to nls()* for the 2021 Google Summer of Code we wanted:

- to provide a packaged version of `nls()` (call it `nlsalt`) coded entirely in R 
  that matches the features in base R or what is packaged in `nlspkg` as described 
  in @PkgFromRbase22. The R version can either serve as the operational code or
  as a reference version to facilitate the maintenance of C, Fortran, or other 
  versions of the software.

- to streamline the overall `nls()` infrastructure. By this we
  mean a re-factoring of the routines so they are better suited to maintenance of both
  the existing `nls()` methods and features as well as new features we or others
  would like to add.

- to explain what we do, either in comments, examples, or separate maintainer documentation.
  Since we are complaining about the lack of explanatory material for the current
  code, we feel it is incumbent on us to provide such material for our own work, and if
  possible for the existing code.

These goals echo themes in the recent discussion by @Vidoni21 and accompanying 
commentary.

## Rationalization of formula specifications

We have seen that
`nls()` uses a different formula specification from the default if the 
`plinear` algorithm is used. This is unfortunate, since the user cannot then simply
add `algorithm="plinear"` to the standard call, making errors more likely.

Nonlinear models, apart from those using the "plinear" algorithm, need all the parameters specified.
That is, the model formula should be written down as it would be computed.
Thus a linear model (which is valid input for nonlinear estimation) should
be specified `y ~ a * x + b` rather than `y ~ x`. 
Moreover, we want to avoid the partially linear parameters appearing 
in the result object as `.lin1`, `.lin2`, etc., where the actual position 
of parameters in the model is not explicit. We are unclear why these parameters
begin with a dot, that is, that they have the "hidden" indicator.

A possible workaround to allow consistency would be to specify the 
partially linear parameters when the `plinear` option is specified. For example, 
we could use `algorithm="plinear(Asym)"` while keeping the general model formula 
from the default specification. This would allow 
for the output of different `algorithm` options to be consistent. However, we have not
yet tried to code such a change, and welcome discussion from those working with
other packages using model formulas.

A further complication of model specification in R is that some formulas require
the user to surround a term with `I()` to inhibit interpretation of arithmetic 
operators. We would prefer that formulas be usable without this complication.

<!-- It is also of interest to consider code that will detect partial linearity,  -->
<!-- rather than expecting the user to provide the special structure and settings. Work in this -->
<!-- area appears to be quite limited, with @Birkisson2013 seemingly a sole contribution -->
<!-- we can find. -->
<!-- Unfortunately, the mention of a "Linear and Nonlinear Discoverer" in @Zhang2011LinNonlin  -->
<!-- is about a class of models that have almost nothing in common -->
<!-- with the present issue. The algorithm in @Birkisson2013 will need investigating further, -->
<!-- but we suspect that it is complicated enough that it will not easily be adapted to  -->
<!-- a program like `nls()`. Possibly a simpler algorithm may be found. -->

<!-- ``` -->
<!-- y ~ x + z + x*z -->
<!-- y ~ a + b*x +c*z + d * (x*z)  # <-- fully specified -->
<!-- y ~ a*Fn1(b, c, x, z)/Fn2() -->

<!-- ``` -->

## Rationalization of indexed models

Indexed models clearly have a place in some areas of research. 
<!-- We need to be able to process models, as the example above, where the -->
<!-- model formula is  -->
We need to be able to process models such as 
`Length ~ a[Strip] + b[Strip] * exp(-Conc / th)`, where `Strip` is the index. Data
`Length` and `Conc` are available for all observations and parameter `th` applies
to every fitted point, but `a[Strip]` and `b[Strip]` apply only to data indexed, which
could be an experimental run number, or some other categorization.

Given that `[ ]` provides the R method to index arrays, the presence of square
brackets in a model
formula signals indexing. The question is how to perform the correct calculations.

<!-- While tedious (??maybe dummies or fastdummies packages make it less so?),  -->
<!-- we could set up the model as a sum of similar models with dummy -->
<!-- variables to turn "on" or "off" the particular case, transforming the model -->
<!-- parameters to simple, non-indexed ones (as does `nls()`). -->

Our view is that inclusion of the indexing **within** `nls()` or similar tools introduces 
complexity that could be avoided by a suitable wrapper function. The modeling
formula must be expanded to encompass all the index cases, likely with indexed
parameters renamed e.g., `a[2]` becomes `a2` as at present with `nls()`. However,
a wrapper could process output to reconstitute the indexing of parameters.

## Streamlining code

We are pessimistic that the overall structure of `nls()` can be streamlined due
to the entanglement of so many features with the complicated mix of R, C and
Fortran. Indeed, despite digging into the code over some months, we do not feel
confident that we sufficiently understand it nor that we could maintain it. On the other
hand, we do believe equivalent functionality can be built, but with at least
some differences in how the features will be accessed.

## Tests and use-case examples

Maintainers of packages need suitable tests and use-case examples in order:

- to ensure packages work properly, in particular, giving
  results comparable to or better than the functions they are to replace;
- to test individual solver functions to ensure they work across the range of calling 
  mechanisms, that is, different ways of supplying inputs to the solver(s);
- to pose "silly" inputs to see if these bad inputs are caught by the programs.

Such goals align with the aims of **unit testing** 
(e.g., *https://towardsdatascience.com/unit-testing-in-r-68ab9cc8d211*, @HWtestthat11, @HWdevtools21
and the conventional R package testing tools).

## A testing package: NLSCompare

One of us has developed a working prototype package at *https://github.com/ArkaB-DS/nlsCompare* .
A primary design objective of this is to allow the summarisation of multiple tests in
a compact output. The prototype has a vignette to illustrate its use.

<!-- Arkajyoti -- do you want to expand? -->
<!-- No. I think it would be tangential to the goal of the paper. We can have a -->
<!-- separate article for that, maybe? -->
<!-- Trying to trim a lot of words, so am making reference to package. Is it ready -->
<!-- for others to use? -->

<!-- When we have a "new" or trial solver function, we would like to know if it gives  -->
<!-- acceptable results on a range of sample problems of different types, starting -->
<!-- parameters, input conditions, constraints, subsets, weights or other settings.  -->
<!-- Ideally we want to be able to get a summary that is easy to read and assess. For -->
<!-- example, one approach would be to list the names of a set of tests with a red, green -->
<!-- or yellow dot beside the name for FAILURE, SUCCESS, or "NOT APPLICABLE". In the last -->
<!-- category would be a problem with constraints that the solver is not designed to  -->
<!-- handle.  -->

<!-- To accomplish this, we need a suitable "runner" program that can be supplied with -->
<!-- the name of a solver or solvers and a list of test problem cases. Problems generally -->
<!-- have a base setup -- a specification of the function to fit as a formula, some data -->
<!-- and a default starting set of parameters. Other cases can be created by imposing -->
<!-- bounds or mask constraints, subsets of the data, and different starts.  -->

<!-- How to set up this "runner" and its supporting infrastructure is non-trivial. While -->
<!-- the pieces are not as complicated as the inter-related parts of the solvers, especially -->
<!-- `nls()`, the categorization of tests, their documentation, and the structuring to make -->
<!-- running them straightforward and easy requires much attention to detail. -->

<!-- Some considerations for our test scripts: -->

<!-- - Is it useful to have a "base" script for each family of test problem, with numbered particular -->
<!--   cases? That is, if we run the scripts in order, we can avoid some duplication of code and -->
<!--   data. -->

<!-- - While we have developed some tags to document the test problem families and cases, we believe -->
<!--   that such tags (essentially summary documentation) will continue to need revision as different -->
<!--   tools and problems are included in scope of `nlsCompare`. -->

<!-- - Similarly, we expect that there will be ongoing review of the structure of the result files. -->

<!-- # Progress so far -->

<!-- Much of our work is reported in https://github.com/nashjc/RNonlinearLS . -->

<!-- ## Reports and articles -->

<!-- - The present article is the central report of the Google Summer or Code 2021 project -->
<!--   "Improvements to nls()". -->

<!-- - VarietyInNonlinearLeastSquaresCodes.Rmd: a review of the different algorithms and -->
<!--   the many choices in their implementation for nonlinear least squares. This is still -->
<!--   a DRAFT at 2022-6-20. -->

<!-- Response to query about TestsDoc -->
<!-- 4. Testing infrastructure, I think, is a general topic and not -->
<!-- particularly relevant to nls(); i.e., it is important for all packages. -->
<!-- So, I don't think we need to include that in RefactoringNLS. Could be a -->
<!-- separate blog maybe?   -->

<!-- Arkajyoti: I think TestsDoc is too preliminary to mention, though the links might be -->
<!-- useful. Perhaps incorporate in the nlsCompare repo. -->
<!-- - TestsDoc.Rmd (?citation) is a survey of testing tools in R. It has more  -->
<!--   general possibilities  -->
<!--   and fits into the subject of regression testing, in which case a  -->
<!--   more extensive literature review will be needed.  Note that this -->
<!--   document reflects the work in the the "Problem sets and test infrastructure" -->
<!--   below. -->

  <!-- need to incorporate MachID and -->
  <!-- the many tests. Arkajyoti: should we think of working this up into a paper for -->
  <!-- JSS or the R-Journal? My  -->
  <!-- view is that this COULD be a long term side-interest for your academic work, but -->
  <!-- that would depend on your own interests as well as opportunities. -->

<!-- This needs more focus of where it may appear. I'm thinking of a blog post -- JN-->
<!-- - DerivsNLS.Rmd ?citation: a document to explain different ways in which Jacobian information -->
<!--   is supplied to nonlinear least squares computation in R. File `ExDerivs.R` is -->
<!--   a DRAFT of a script to provide examples. -->

## Documentation and resources

In our investigation, we built several resources, which are now part of 
the repository *https://github.com/nashjc/RNonlinearLS/*. Particular items
are:

- A BibTex bibliography for use with all documents in this project, but which 
  has wider application to nonlinear least squares projects in general
  (*https://github.com/nashjc/RNonlinearLS/blob/main/BibSupport/ImproveNLS.bib*).

- `MachID.R` offers a suggested concise summary function
  to identify a particular computational system used for tests. A discussion 
  of how it was built and the resources needed across platforms is given in
  at *https://github.com/nashjc/RNonlinearLS/tree/main/MachineSummary*.

- @PkgFromRbase22 is an explanation of the construction of the `nlspkg` from the
  `nls()` code in R-base.

- As the 2021 Summer of Code period was ending, one of us (JN) was invited to prepare
  a review of optimization in R. Ideas from the present work have been instrumental
  in the creation of @NashWires22.

<!-- ?? probably don't need to include -->
<!-- - WorkingDocument4ImproveNLS.Rmd: essentially a way to record what we have worked on. -->
<!--   A project diary. -->

<!-- ?? does not seem to add enough? -->
<!-- ## Problem sets and test infrastructure -->

<!-- ?? need to clean up -->

<!-- We have several test problems and variants thereof in the `inst/scripts/` directory of -->
<!-- the `nlsCompare` package available on Github (https://github.com/ArkaB-DS/nlsCompare). -->
<!-- We direct the reader to that package for documentation of the test infrastructure,  -->
<!-- in particular the problems and methods files (`problems.csv` and `methods.csv`) and -->
<!-- the various functions invoked by `run.R` to produce an output file in CSV form also. -->

<!-- Towards the end of the project, we have focused our attention on the `nlsCompare` -->
<!-- package, which looks at evaluating and comparing functions for nonlinear least -->
<!-- squares problems. We use many of the same tests as checks that are or will be  -->
<!-- built into our packages for such problems e.g., `nlsalt`. These use the `testthat` -->
<!-- structure and may include verification of outputs that are specific to a package, -->
<!-- such as the upper triangular matrix R of the QR decomposition that has been  -->
<!-- computed for a Jacobian. Since such an object will not be computed by all methods,  -->
<!-- testing it in `nlsCompare` makes no sense, and in that package we concentrate on -->
<!-- the minimum sum of squares and the associated model parameters. -->

<!-- ## Code development -->

<!-- A summary of our main results: -->

<!-- - nlspkg: a packaged version of the `nls() code from R-base. Thanks to Duncan Murdoch -->

<!-- - nlsalt: attempt to mirror `nls() behaviour entirely in R. This is UNFINISHED.  -->
<!--   The effort showed that the structure of the programs was difficult to follow,  -->
<!--   undocumented, and unsuited to adding improvements like the Marquardt stabilization.  -->
<!--   We were able to get a pure-R version of `numericDeriv()` and rework most of the functions  -->
<!--   of `nlsModel` (but not `nlsModel.plinear`). This work may continue after the project -->
<!--   formally ends, but collaboration is likely needed with workers who have a deep knowledge -->
<!--   of R internals. `nls-flowchart.txt` was a start at documentation of the structure of `nls()`. -->

<!-- <!-- - nls-changes-for-small-residuals-in-nls-R-4.0.2.zip: collected material for the  --> 
<!-- <!--   fix by JN to the relative offset convergence criterion failure when there are small  --> 
<!-- <!--   residuals in problems sent to ``nls()`. Now part of R. -->

<!-- - the safeguarded relative offset convergence criterion that allows small-residual problems -->
<!--   to ba handled with `nls()` is now in the R base code. -->

<!-- - `nlsj` is a refactoring of some `nls()` functionality. ?? What are key features?? -->
<!--   This is an interim package for exploration of ideas and will NOT, as it currently -->
<!--   is presented, become a distributed package. -->

<!-- - nlsralt: this is intended to be a modified version of Nash and Murdoch package  -->
<!--   `nlsr` to examine possible improvements discovered as a result of this project.  -->
<!--   This is EXPERIMENTAL. -->

<!-- For example,  -->
<!-- `nlsr` does not currently use the `subset` argument correctly. We also need to document -->
<!-- changes e.g., to numerical derivatives. However, these are tasks for JN rather than AB. -->

# Strategic choices in nonlinear model estimation

It is possible that `nls()` will remain more or less as it has been for the past
two decades. Given its importance to R, the focus of discussion should be the measures
needed to secure its continued operation for legacy purposes and how that may be approached.
We welcome an opportunity to participate in such conversations.

To advance the stability and maintainability of R, we believe the program objects (R functions)
that are created by tools such as `nls()` should have minimal cross-linkages and side-effects.
The aspects of `nls()` that have given us the most trouble:

- The functions that compute the residuals and jacobians often presume
  the data and parameters needed are available in a particular environment. As long
  as the correct environment is used, this provides a compact syntax to invoke the
  calculations. The danger is that the wrong data is accessed if the internal search finds a
  valid name that is not the object we want.

- Weights, subsets, and various contextual controls (such as that for the `na.action` option)
  are similarly taken from the first available source, making for a very simple
  invocation of calculations, but the context is hidden from the user. Furthermore, it can be
  difficult for those of us trying to maintain or improve the code to be certain we have the
  context correct.

- The mixing of R, C and Fortran code was necessary for computational performance in the past.
  Lacking decent developer documentation, programmers now face a lot of work to fix or improve code. 
  We believe in having at least a working reference version of code in a single programming language. 
  If necessary, by
  measuring ("profiling") the code, we can find bottlenecks and substitute for those slower 
  parts of the reference code.

- A design that isolates the setup, solution, and post-solution parts of
  for complicated calculations reduces the number of objects that must be kept in 
  alignment. On the other hand, it may mean users have to add more commands to complete
  their computations.

- All iterative codes should return the best solution they have found so far, even if there
  are untoward conditions reached, such as a singular Jacobian. This modification could be
  made to `nls()` in the short term, correcting the issue identified above under "Failure to
  return best result achieved".

- Given the existence of examples of good practices such as analytic derivatives, stabilized 
  solution of Gauss-Newton equations and bounds-constrained parameters, base R tools should 
  be moving to incorporate them.

We anticipate that tools like `nlsr` will become more attractive as users discover
the increased robustness and general capabilities, along with the more recent documentation.
From a design point of view, a key difference in approach between `nls()` and 
`nlsr::nlxb()` is that `nls()` 
builds a large infrastructure, especially the `nlsModel()` function,  from which the Gauss-Newton
iteration can be executed and other statistical information such as profiles can be
computed, while `nlxb()` returns quite limited information, and 
computes what is needed on an "as and when" basis. This follows a path that one of us
(JN) established almost 50 years ago with the software that became @cnm79, separating
setup, solver, and post-solution analysis phases of computations. There is, however,
the wrapper `wrapnlsr()` that solves the nonlinear estimation problem with the solver
of `nlsr` then calls `nls()` to build the object of class `nls` for which profile log-likelihood
information can be computed.
<!-- Here, that last  -->
<!-- phase is not part of `nlxb()`, but is the domain of other functions in the `nlsr` -->
<!-- package. -->

<!-- The `nls()` approach, as implemented in the base R code and `nlspkg` leads to considerable -->
<!-- entangling of the different functions and capabilities. Even after -->
<!-- both the preparation for and carrying out of the present project, we do not feel confident -->
<!-- to explain the code completely, nor to maintain it. However, we have made some  -->
<!-- forays into render parts of the code in R, and to understand where difficulties -->
<!-- lie. -->

<!-- ??fix next bits -->
<!-- - The different dialects of modeling expressions as detailed above should be rationalized. This -->
<!--   would, however, be a painful and mainly political activity. -->

<!-- - The tools for analytic derivatives that exist in parts of `nls()` for self-starting models and -->
<!--   in `nlsr` for problems specified as models that use expressions for which derivatives exist  -->
<!--   should be more generally available, but the task is one of tedium and limited reward to the -->
<!--   persons who do it, while benefiting the R community substantially. -->

<!-- - Similarly, default methods need stabilized solvers e.g., Marquardt methods, though to date -->
<!--   a mechanism to incorporate such modifications in `nls()` has eluded us. -->

<!-- - Bounds constraints are sufficiently important that they should be part of the default -->
<!--   methods -->
<!--   (rather than algorithm = "port" for `nls()`). `minpack.lm` should handle them correctly. -->

# Acknowledgement

Hans Werner Borchers has been helpful in developing the proposal for this project
and in comments on this and related work. Heather Turner co-mentored the project 
and helped guide the progress of the work. Exchanges with Fernando Miguez helped to
clarify aspects of `selfStart` models and improve package `nlsr` in that respect.
Colin Gillespie (package `benchmarkme`) has been helpful in guiding our attempts
to succinctly summarize computing environments.
